---
title: "Introduction to POSSA. Sequential tests for two independent means (t-tests)."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library('POSSA')
```

Below, the essence of `POSSA`'s power (and type 1 error rate, T1ER) calculation for sequential designs is explained using the simple example of a t-test. Of course, for t-tests specifically, you could use other and potentially easier software – however, the point here is to understand how the `POSSA` works, and, subsequently, it should be possible for you to extend this example to virtually any statistical null hypothesis test. Cases of correlated samples and multiple hypotheses are explained on separate pages, but these too are straightforward extensions of the present example.

All parameters mentioned below are described in detail in full the documentation (see `?sim` and `?pow`).

So let's say you want to compare the means of two independent continuous variables with a simple t-test. Your smallest effect size of interest (SESOI) is `5` units (arbitrary units that could represent anything – e.g., the height of a plant in cm, or the time of performing a task in seconds, etc.).

### User functions for sample simulation and testing

First, write a function that simulates the samples (i.e., observed values) for a single instance of the hypothetical experiment in case the null hypothesis is true (no actual difference between the two population means) as well as in case the alternative hypothesis is true (there is in fact difference). 

If the null hypothesis "H0" is true, the two samples have the same mean, let's say `0` units; and let's assume normally distributed values an SD of `10` units for each sample. For this, two samples can each be simulated via the function `rnorm(sampleSize, mean = 0, sd = 10)` (where `sampleSize` is a variable to be provided via another function, as explained later). You can assign this to `sample1`, which represents a baseline variable (e.g., "group A"), and, separately, to `sample2_h0`, which represents a sample that does not differ from the baseline ("H0" true). If the null hypothesis "H1" is true, the baseline sample may again have a mean of `0`, but the other sample should have the SESOI, `5` units (again assuming an SD of `10`) – which can be simulated as `rnorm(sampleSize, mean = 0, sd = 10)`. This can be assigned to `sample2_h1`, which represents a sample that does differ from the baseline ("H1" true). (The baseline sample has identical properties in case of H0 and H1, so for it's enough to simulate one.)

All this is to be put into a function, which takes a `sampleSize` argument and returns the simulated samples as a named list, as follows.

```{r}
customSample = function(sampleSize) {
  list(
    sample1 = rnorm(sampleSize, mean = 0, sd = 10),
    sample2_h0 = rnorm(sampleSize, mean = 0, sd = 10),
    sample2_h1 = rnorm(sampleSize, mean = 5, sd = 10)
  )
}
```

Note that the `customSample` and `sampleSize` variables were written with camel case (likeThis), while the `_h0` and `_h1` endings used underscores (like_this). This alternate notation is used in the present tutorial to help distinguish what names can be freely chosen, and what is necessarily named so for the `POSSA` package: custom names are all camel case, while underscores indicate notation necessary for `POSSA`. Here specifically, the list names for the sample that varies depending on whether the null (H0) and alternative (H1) hypothesis is true must be indicated with `_h0` and `_h1` endings, respectively, with a common root (here: `sample2`). The root name or the other variable names (`customSample` and `sampleSize`) could be named differently, and everything would work just as well.

Now, write a function that performs the test; here, a one-sided t-test (with equal variances assumed – this is just for the example's outcome's comparability for other software, normally Welch's test should be preferred).

The function's parameters must be identical to the names of the list elements returned by the sample function (here: `sample1`, `sample2_h0`, and `sample2_h1`). The returned value must be a named vector that contains the derived p values. The name of each p value must start with `p_`, and end with `_h0` for the "H0" outcome and with `_h1` for the "H1" outcome. (Below, it's simply `p_h0`/`p_h1`, but it could just as well be, e.g., `p_my_t.test_h0`/`p_my_t.test_h1`).

```{r}
customTest = function(sample1, sample2_h0, sample2_h1) {
  c(
    p_h0 = t.test(sample1, sample2_h0, 'less', var.equal = TRUE)$p.value,
    p_h1 = t.test(sample1, sample2_h1, 'less', var.equal = TRUE)$p.value
  )
}
```

To make sure the two functions are working correctly, as a quick test you can run the following line (with any applicable number argument).

```{r}
do.call(customTest, customSample(55))
```

This passes `55` as (arbitrary) sample size to `customSample`, which passes the created samples to `customTest`, which finally returns the named vector with the `p_h0` and `p_h1` p values. (Given the randomization, if you run this line repeatedly, the outcome should differ each time – however, with large enough numbers, `p_h1` is likely to be noticeably smaller then `p_h0`.)

### POSSA's *sim* and *pow*

Now, the `POSSA::sim` function can be used with `customSample` for `fun_obs` (function for observation values) and `customTest` for `fun_test` (function for statistical testing), and any desired numbers of observation. (This runs the simulation `45000` times by default, so it takes a while. For quick testing, set the number to a lower value via the `n_iter` parameter, e.g., `n_iter = 500`.)

```r
df_ps = sim(fun_obs = customSample,
            n_obs = 80,
            fun_test = customTest)
```

The returned `df_ps` contains all the simulated p values, and can be directly passed to `POSSA::pow` to get the power for any specified alpha level.

```r
pow(df_ps)
```
This prints to the console the following.

```r
#> # POSSA pow() results #
#> N(average-total) = 160.0 (if H0 true) or 160.0 (if H1 true)
#> (p) Type I error: .05140; Power: .93327
#> Local alphas: (1) .05000
```

This is a fixed sample design with 80 observations per each of the two groups. The function uses an alpha of `.05` by default, this may be modified e.g. to `.01` below.

```r
pow(df_ps, alpha_global = 0.01)

#> # POSSA pow() results #
#> N(average-total) = 160.0 (if H0 true) or 160.0 (if H1 true)
#> (p) Type I error: .00991; Power: .79069
#> Local alphas: (1) .01000
```

So far, we can easily get the same information from other software, e.g. `pwr::pwr.t.test(n = 80, d = 0.5, alternative = 'greater')` (returns `power = 0.93368`) and `pwr::pwr.t.test(n = 80, d = 0.5, sig.level = 0.01, alternative = 'greater')` (returns `power = 0.79068`).

However, sequential designs are extremely easy to be expanded from the above `sim` and `pow` examples. One just has to specify multiple instances of observation numbers for the `n_obs` in `sim` and specify a local alphas in `pow`, e.g. as below.

```r
df_ps_seq = sim(fun_obs = customSample,
                n_obs = c(27, 54, 81),
                fun_test = customTest)
pow(df_ps_seq, alpha_locals = NA)

#> # POSSA pow() results #
#> N(average-total) = 158.6 (if H0 true) or 98.8 (if H1 true)
#> (p) Type I error: .05000; Power: .89922
#> Local alphas: (1) .02288; (2) .02288; (3) .02288
#> Likelihoods of significance if H0 true: (1) .02296; (2) .01631; (3) .01073
#> Likelihoods of significance if H1 true: (1) .42324; (2) .32298; (3) .15300
```

The `alpha_locals = NA` input specifies that all local alphas are `NA`, in which case they will all gain a single identical number (corresponding to Pocock's correction in case of equally spaced looks). Rounding to the conventional 3 fractional digits, the adjusted constant local alpha number corresponds to the exact statistics (`.023`) provided in other software, e.g. `gsDesign::gsDesign(k = 3, test.type = 1, alpha = 0.05, sfu = 'Pocock')`.

The console output contains only the most important information, but, when assigned, the `pow` function returns a data frame with detailed information per each look (and their totals), including the specified sample sizes, the numbers of iterations, etc. The printed information includes

- the average number of subjects in H0 and in H1 scenarios (i.e., taking of all simulations, on average at how many participants the experiment was stopped),
- the T1ER (here, it's for a single p value, but, in case of multiple p values, it would be a combined T1ER as specified)
- the power (again, would be combined in case of multiple p values), 
- the local alphas per look (and per each p value; but in the present case there is only one, named simply `p`, the root derived from `p_h0`/`p_h1`),
- the ratios of significant findings per each look (corresponding to the local alphas).

The look numbers are always in parenthesis, such that "`(1)`" designates the first look, "`(2)`" the second look, etc.

### Adjusting local alphas

`POSSA` helps obtain the local alphas, in sequential designs, that result in a specified overall (global) T1ER, which may also be called "global alpha". This procedure is automatized for most practical cases, but still users should ideally have a basic idea of how this works. However, impatient readers (who do not want to have customized adjustments) can skip to the **Specifying initial local alphas** section below.

The essential mechanism is a staircase procedure, where the local alphas are continually increased when the T1ER is smaller than specified and decreased when the T1ER is larger then specified, and each direction change (from decrease to increase or vice versa) moves onto a smaller step size in the adjustment. For example, the simplest case is when all local alphas are provided as `NA`, as above: in this case, the initial replacement value (which may be modified via the `adj_init` parameter) is by default the global alpha divided by the maximum number of looks, as a rough initial estimation. Subsequently, the given p values are tested with this setting, and a global T1ER is calculated. If it is larger than specified, the replacement value is decreased, or vice versa. The amount of decrease is the first step, by default `0.01`. Then the testing is repeated, and the change is repeated in the same direction until the obtained T1ER passes the specified T1ER (e.g., becomes smaller, when initially larger). Then the replacement value is increased with a smaller second step size, by default `0.05`. And so on, until either there are no more steps (as specified) or (more typically) the obtained T1ER is close enough to the specified one (e.g., by default, matches it up to 5 fractional digits; may be specified via `alpha_precision`), at which point the procedure stops and the obtained local alphas (each having the same value) and the corresponding results are returned and printed.

The adjustment actually happens via a function that can be specified via the `adjust` parameter, but, by default, whenever there is at least one `NA` value among the given `alpha_locals` argument, the function implements the above-described procedure, and looks as follows.

```r
adjust = function(adj, prev, orig) {
    prev[is.na(orig)] = adj
    return(prev)
}
```

In this function, `adj` means the adjustment value (in this case a replacement value), `prev` the previous vector of local alphas (obtained in the last adjustment), and `orig` the original vector of local alphas as provided for `alpha_locals`. (For the first adjustment, `prev` is the same as `orig`). Any sort of custom function may be provided as `adjust`, but it cannot contain any parameter other than these three (`adj`, `prev`, and `orig`).

By default, if no `NA` value is found among the given `alpha_locals`, the `adjust` function uses multiplication to adjust the given values (with `adj_init` set to `1`), as follows.

```r
adjust = function(orig, adj) {
    return(orig * adj)
}
```

By default, there are altogether 11 step sizes, chosen in a way that typically results in 5-fractional-digit T1ER match before running getting to the last step, and the calculation typically takes only a few seconds. (Step sizes can be modified via the `staircase_steps` parameter.)

### Specifying initial local alphas

Given the above-described default mechanisms, there are, without modifying the adjustment procedure, two easy ways to provide the argument for `alpha_locals` (for any given p value): (a) a vector that includes (or consist entirely of) `NA` values that are to be replaced with a constant number that results in the desired global T1ER, and (b) a vector with all numeric values that are each to be multiplied with a number in order to obtain a vector of alphas that again result in the desired global T1ER.

However, you can also simply check the results, without any adjustment, of any given set of local alphas, by setting `adjust = FALSE`.

Here are some examples, all using the data frame "`df_ps_seq`" created above with `POSSA::sim()`.

To use Haybittle–Peto boundary (all interim local alphas `.001`), you can specify `.001` for each local alpha except for the last, which could be `NA` to be adjusted for the given global alpha (here: `.025`), as below.

```r
pow(df_ps_seq, alpha_locals = c(0.001, 0.001, NA), alpha_global = 0.025)

#> # POSSA pow() results #
#> N(average-total) = 161.8 (if H0 true) or 140.4 (if H1 true)
#> (p) Type I error: .02500; Power: .88411
#> Local alphas: (1) .00100; (2) .00100; (3) .02431
#> Likelihoods of significance if H0 true: (1) .00111; (2) .00067; (3) .02322
#> Likelihoods of significance if H1 true: (1) .09224; (2) .21596; (3) .57591
```

In this specific scenario, despite noticably decreased average sample in case of H1 (`140.4` instead of the fixed `162`), the adverse effect of interim stops on T1ER is so minimal (see, in case of H0, the tiny ratio significant [false positive] findings at the first and second looks), that the last look's alpha (`.02431`) hardly differs from the specified global alpha (`.025`).

You can also check what happens without adjustment, using `adjust = FALSE`.

```r
pow(df_ps_seq, alpha_locals = c(0.001, 0.001, 0.025), alpha_global = 0.025, adjust = FALSE)

#> # POSSA pow() results #
#> N(average-total) = 161.8 (if H0 true) or 140.4 (if H1 true)
#> (p) Type I error: .02551; Power: .88618
#> Local alphas: (1) .00100; (2) .00100; (3) .02500
#> Likelihoods of significance if H0 true: (1) .00111; (2) .00067; (3) .02373
#> Likelihoods of significance if H1 true: (1) .09224; (2) .21596; (3) .57798
```

Which leads to the same conclusion from the reverse perspective: the T1ER hardly exceeds `.25`.

Nonetheless, more substantial sample reduction can be achieved with somewhat more liberal interim local alphas; besides, normally it makes more sense to increase the local alphas gradually (because, in brief, given the greater uncertainty of the initial smaller samples, beginning with lower local alphas and gradually increasing them provides the most efficient "information gain" for the same eventual T1ER). For this, one could base the adjustment on one of the popular boundary methods used for parametric tests, such as the O'Brien-Fleming bounds. These bounds for parametric tests can be easily calculated via other packages. These bounds may not apply to nonparametric tests in exactly the same way, but the principle of exponentially increasing local alphas remains the same. (To decide what fits your specific scenario best, you could also check the outcomes of various versions of the Wang-Tsiatis bounds by changing its _Delta_ parameter; e.g., in `gsDesign::gsDesign()`, set `sfu = 'WT'` for Wang-Tsiatis, and vary its Delta via `sfupar`.) Most importantly in any case, the adjustment of alphas in `POSSA` will ensure that the T1ER is at the specified level. This will only make a difference for other custom tests and multiple hypotheses, but not for the present t-test example.

For our example of three equally spaced interim looks, O'Brien-Fleming bounds can be calculated e.g. via `gsDesign::gsDesign(k = 3, test.type = 1, alpha = 0.05, sfu = 'OF')`, giving the local alphas of `0.0015`, `0.0181`, and `0.0437`. (The same is returned by `rpact::getDesignGroupSequential(typeOfDesign = "OF", informationRates = c(1/3, 2/3, 1), alpha = 0.05)`.)

To merely calculate the T1ER and power for any specific tests, again use `adjust = FALSE`.

```r
pow(df_ps_seq, alpha_locals = c(0.0015, 0.0181, 0.0437), alpha_global = 0.05, adjust = FALSE)

#> # POSSA pow() results #
#> N(average-total) = 160.8 (if H0 true) or 118.7 (if H1 true)
#> (p) Type I error: .04951; Power: .93087
#> Local alphas: (1) .00150; (2) .01810; (3) .04370
#> Likelihoods of significance if H0 true: (1) .00162; (2) .01809; (3) .02980
#> Likelihoods of significance if H1 true: (1) .11531; (2) .57211; (3) .24344
```

(To verify the power via other software, see e.g. `summary(rpact::getSampleSizeMeans(rpact::getDesignGroupSequential(typeOfDesign = "OF", informationRates = c(1/3, 2/3, 1), alpha = 0.05, beta = 1-0.93087), alternative = 0.5))`.)

Since these local alphas were specifically chosen to have a T1ER of `.05` for the t-test comparison used here, they will hardly change even with `adjust = TRUE`. However, to test the adjustment function, we can set a different global alpha, e.g. `.01`, so that the adjustment will reduce each local alpha value by multiplication.

```r
pow(df_ps_seq, alpha_locals = c(0.0015, 0.0181, 0.0437), alpha_global = 0.025)

#> # POSSA pow() results #
#> N(average-total) = 161.4 (if H0 true) or 126.9 (if H1 true)
#> (p) Type I error: .02498; Power: .87487
#> Local alphas: (1) .00071; (2) .00862; (3) .02080
#> Likelihoods of significance if H0 true: (1) .00089; (2) .00876; (3) .01533
#> Likelihoods of significance if H1 true: (1) .07629; (2) .49747; (3) .30111
```

(Cf. `summary(rpact::getSampleSizeMeans(rpact::getDesignGroupSequential(typeOfDesign = "asUser", informationRates = c(1/3, 2/3, 1), userAlphaSpending = c(.00071, .0089, .0244), beta = 1-.87487), alternative = 0.5))`.)

Finally, just for quick demonstration, for the same scenario, here is a function that adjusts all values via additions (constant increase/decrease of all local alpha values). Also, here the global alpha is set to be larger (`alpha_global = 0.1`), just so as to make the local alphas increase too in this case.

```r
pow(
  df_ps_seq,
  alpha_locals = c(0.0015, 0.0181, 0.0437),
  alpha_global = 0.1,
  adjust = function(adj, prev, orig) {
    return(orig + adj)
  }
)

#> # POSSA pow() results #
#> N(average-total) = 161.4 (if H0 true) or 126.9 (if H1 true)
#> (p) Type I error: .02498; Power: .87487
#> Local alphas: (1) .00071; (2) .00862; (3) .02080
#> Likelihoods of significance if H0 true: (1) .00089; (2) .00876; (3) .01533
#> Likelihoods of significance if H1 true: (1) .07629; (2) .49747; (3) .30111
```

### Futility bounds

By default, the interim local alphas will all be zero (i.e., no stopping for significance), and the final local alpha will equal the given global alpha – meaning a simple fixed design (with the maximum specified observation number).

```r
pow(df_ps_seq)

#> # POSSA pow() results #
#> N(average-total) = 162.0 (if H0 true) or 162.0 (if H1 true)
#> (p) Type I error: .04860; Power: .93636
#> Local alphas: (1) none; (2) none; (3) .05000
#> Likelihoods of significance if H0 true: (1) 0; (2) 0; (3) .04860
#> Likelihoods of significance if H1 true: (1) 0; (2) 0; (3) .93636
```

This is to make the setting of _futility bounds_ alone easy and straightforward.

pow(df_ps_seq, fut_locals = 0.5)


-- To skip, at any look, alpha/futility stop, just set to 1/0.

### Arbitrary stops

In a real experiment, you may not know in advance the exact number of observations at the stopping point. For example, you plan (and preregister) to open an initial 30 slots for participation, after which you would perform the first testing (first interim "look") – however, some participants will probably have to be excluded (e.g. because of failing attention checks), expected to be only a few, but cannot be known exactly in advance.

Assuming that minor variations in power do not matter, you can simply calculate with the approximate observation numbers (at the given planned looks) in advance, and, when the actual observation numbers are known, the calculation is repeated with these latter numbers to ensure that the T1ER remains as desired. For example, taking into account some necessary exclusions, the expected observation numbers with two interim looks are `27` (first look), `54` , and `81`. This may be calculated as before, using O'Brien-Fleming bounds (as provided by, e.g., `gsDesign::gsDesign(test.type = 1, alpha = 0.05, sfu = 'OF', timing = c(27, 54, 81)/81)`, then adjusted by `POSSA`).

```r
df_ps_planned = sim(fun_obs = customSample,
                n_obs = c(27, 54, 81),
                fun_test = customTest)

pow(df_ps_planned, alpha_locals = c(0.0015, 0.0181, 0.0437), alpha_global = 0.05)

#> # POSSA pow() results #
#> N(average-total) = 160.8 (if H0 true) or 118.5 (if H1 true)
#> (p) Type I error: .05000; Power: .93164
#> Local alphas: (1) .00152; (2) .01830; (3) .04419
#> Likelihoods of significance if H0 true: (1) .00164; (2) .01831; (3) .03004
#> Likelihoods of significance if H1 true: (1) .11604; (2) .57309; (3) .24251
```

Now, let's say that, due to somewhat more exclusions than expected, the initial sample contains only 24 valid observations. You can rerun the simulation with the modified first observation number, and run the power calculation with initial alphas corrected with `gsDesign::gsDesign(test.type = 1, alpha = 0.05, sfu = 'OF', timing = c(24, 54, 81)/81)`.

```r
df_ps_first = sim(fun_obs = customSample,
                n_obs = c(24, 54, 81),
                fun_test = customTest)

pow(df_ps_first, alpha_locals = c(0.0009, 0.0182, 0.0438), alpha_global = 0.05)

#> # POSSA pow() results #
#> N(average-total) = 161.0 (if H0 true) or 120.6 (if H1 true)
#> (p) Type I error: .05000; Power: .93484
#> Local alphas: (1) .00091; (2) .01850; (3) .04453
#> Likelihoods of significance if H0 true: (1) .00087; (2) .01760; (3) .03153
#> Likelihoods of significance if H1 true: (1) .07056; (2) .61742; (3) .24687
```

Let's say that the p value was not below the local alpha (`.02295`), so you move open another 30 slots, and once again end up with more exclusions than expected, resulting in only `49` valid observations. The adjusted calculation is then as follows.

```r
gsDesign::gsDesign(test.type = 1, alpha = 0.05, sfu = 'OF', timing = c(24, 49, 81)/81)
# gives new initial local alphas

df_ps_second = sim(fun_obs = customSample,
                   n_obs = c(24, 49, 81),
                   fun_test = customTest)

pow(df_ps_second, alpha_locals = c(0.0009, 0.0145, 0.0447), alpha_global = 0.05)

#> # POSSA pow() results #
#> N(average-total) = 161.0 (if H0 true) or 119.6 (if H1 true)
#> (p) Type I error: .04996; Power: .93153
#> Local alphas: (1) .00090; (2) .01453; (3) .04479
#> Likelihoods of significance if H0 true: (1) .00091; (2) .01360; (3) .03544
#> Likelihoods of significance if H1 true: (1) .07009; (2) .53733; (3) .32411
```

(And, if needed, a similar modified recalculation could be done for the last look.)

As can be seen, the power differences are very small. Nonetheless, if it is important to keep the same level of power with high precision (e.g., up to 3 fractional digits), you can plan for (and preregister) to conditionally adjust (increase/decrease) the subsequent interim (if any) and final sample sizes to achieve the same power. (Technically, another option is to conditionally adjust the alpha level, i.e., the T1ER, but that's rather questionable.)
